Probabilistic Programming

### 2014-04-21 paper/paper.tex
Work on abstract and intro

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage[margin=3cm]{geometry}

@@

\author{
	Carlos Diaz
	\and
	Dan Hassin
	\and
	Charles Lehner
	\and
	Dan Viterise
}

@@

\begin{abstract}
	We present Probabilistic Programming, a system for modeling abstract syntax
	trees of programs with Markov chains, which we use to generate new programs.
	We attempt to use this as an assistive device to guide programmers through
	constructing JavaScript programs.
\end{abstract}

@@

\subsection{Motivation}

For many people just starting out, programming can seem like a confusing and
tedious activity. It can be unclear for many new programmers where they should
begin and what should come next while coding. Without any proper guidance, a
novice programmer is left on their own to scour thousands of pages online to
discover what should come next in their code. Then, once a person becomes more
adept at programming, he or she may notice that they are typing out the same or
similar lines of code within each new program. This repetitiveness slows down
the user and can make programming a much more tiresome task than it needs to be.
These observations steered our group towards developing a way for someone to
code with more instruction and productivity. The Probabilistic Programming
method allows for a user to code in JavaScript using a web application. This web
app provides a list of suggestions for the user's next line of code depending on
where the user is in their program. Not only is this helpful for new programmers
that need suggestions for what to write next in their code, this system allows
for expert programmers to write code with more ease and efficiency.

@@

We chose to use a Markov chain to drive the code snippet generation, because
(a) , and (b). A Markov Chain is a stochastic process
that uses the current state and a set of probabilities to determine what state
comes next. In the most simple case of a Markov Chain, the next action is based
only upon the current situation and all other previous actions are forgotten.
Thus, this process is essentially memoryless because all past states can be
forgotten once a new state is reached. A good example of a Markov Chain is a
random walk where at each integer value, the process chooses a new value such
that the two possible values are the current value plus 1 and the current value
minus 1. The transition probabilities for both of these options is .5. This is a
good example of a Markov Chain because the next value on the random walk only
depends on the current value in the walk; all previous values are not needed to
choose the next value \cite{markov}.

@@

and ${Path}({node}_1, {edge}_1) = {node}_2$, ${Path}({node}_2, {edge}_2) = {node}_3$, ..., ${Path}({node}_d, {edge}_d) = {newnode}_k$, where ${newnode}_k$ is randomly selected from ${newnode}_1, ..., {newnode}_n$ given the probabilities \break $P({newnode_1}), ..., P({newnode_n})$. From this it can be seen that generating programs is simply a repetitive application of this function, where for the chosen ${newnode}_k$, we can generate each child of ${newnode}_k$ with the following formula. \begin{align*} &\forall {edge} \in {Features}({newnode}_k), \\
& \hspace{3em} {Path}({newnode}_k, {edge}) = f({node}_2, {edge}_2, ..., {node}_{d}, {edge}_{d}, {newnode}_k, {edge}).
\end{align*}

@@

@@

The Guided IDE offers users an environment for building a program entirely using
our model, using program snippets generated at the cursor's context as building
blocks. The editor presents a window for the program text, and and list of
suggestion code snippets. The user can move a cursor through the code window at
positions corresponding to boundaries of program nodes. As the user moves the
cursor, the snippet suggestions are regenerated using the cursor's node path, to
continuosly provide suggestions relevant to the current context of the program. When the
user selects a suggestion snippet, its AST is inserted into the program's AST at
the cursor's position in the tree, and the code window is redrawn. The user can
move the cursor within newly inserted blocks, so that the program can grow
deeper. The initial corpus for the IDE's model consists of the JavaScript files
that comprise our project.

@@

Instead of having a movable cursor in the code window, have stubs in the
program. Each stub would be a marker in place of an AST node. The user would
select a stub and be given choices of snippets to replace it. Selecting a
snippet would move the current snippet into the program in place of the selected
stub.  The generated snippets could themselves have stubs in them, so that the
user may continue to complete stubs and expand their program.  At the level of
variable identifiers, the snippet suggestions would be taken from the computed
list of in-scope variables at the position of the selected stub, unless the
completion was for a variable declaration, in which case the user would be
prompted to name a new variable.

@@

\section{Conclusions \& Open Problems}

@@

Just as our approach of modeling a program with a tree-based Markov chain is a
higher order model than a linear Markov chain of the program source code tokens,
there may be applications of probabilistic modeling to higher-order
representations of programs than the AST, such as a higher-order abstract syntax
\cite{hosa}. Such an application may model the relationship between different
parts of an AST, and handle scoping of variables to model identifiers
abstractly.

With a more sophisticated model of identifiers that could generate programs with
identifiers in meaningful relation to eachother, Probabilistic Programming could
be used to generate semantically valid programs, for greater assistive,
expressive, and programmatic power.

@@

The source code for Probabilistic Programming is available online under an open
source license:\\
\url{https://github.com/RocHack/jschain}.\\
The web IDE can be found here:
\url{https://csug.rochester.edu/u/clehner/jschain/web/}.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage[margin=3.75cm]{geometry}

@@

\author{
	Carlos Diaz, Dan Hassin, Charles Lehner, Dan Viterise \\
    \vspace{0pt}\\ Department of Computer Science \\University of Rochester
}

@@

\begin{abstract}
	We present Probabilistic Programming, a system for generating JavaScript programs and code snippets based off a corpus of JavaScript source files.
	In this paper, we will demonstrate storing statistical
    data from non-linear structures (abstract syntax trees of programs) in a Markov model,
    and of course using the model to generate new trees. We present a proof of concept for a web-based
    assistive program editor that utilizes this technology to guide programmers through
	constructing JavaScript programs.
\end{abstract}

@@

This paper will begin by discussing Markov chains and how we set out to use them for our goal. We will then describe in detail how we implemented a Markov chain for program syntax trees, how we were able to generate code using it, and finally we'll discuss our results and address the limitations of our product.

@@

A Markov chain is a mathematical system that undergoes a stochastic process whereby
the current state and a set of probabilities are used to determine what state
to transition to next. In the most simple case of a Markov chain, the next action is based
only upon the current state and all other previous actions are forgotten.
Thus, this process is essentially ``memoryless'' because all past states can be
forgotten once a new state is reached \cite{markov}. In this way, Markov chains can be likened
to deterministic finite automata. So while they cannot remember a limitless amount of their history,
calling them ``memoryless'' may be misleading, as they can indeed be implemented to remember a fixed
size of their state history.

Markov chains find application in many different fields. Familiar with Markov chains used for generating random human-sounding text (by training a model based on bigrams and trigrams of common speech), we were interested in perhaps more novel domains. We chose to use a Markov chain to drive code generation, because, like text, we feel that code can be personal and expressive, and the idea of generating code possibilities from a training set seemed like exciting research. It was also challenging -- since randomly generating code tokens would most likely produce syntactically invalid programs, we looked at program abstract syntax trees. And as the research developed, we found ourselves with a new vision of a semifeasable product.

\subsection{Emerging goal}

For many novice programmers, programming can seem like a confusing and
tedious activity. It can be unclear for many new programmers where they should
begin and what should come next while coding. They may conceptually understand
what the next step should be, and yet be unable to express it with code. Or, they
could be unsure of the right next step conceptually, and some code could inspire
them to understand it.

Even once one becomes more
adept at programming, he or she may notice typing the same or
similar lines of code in each new program. This repetitiveness slows down
the user and can make programming more tiresome a task.

The prospect of using a Markov chain to generate code snippets in an assistive, predictive, integrated development environment (IDE) for JavaScript seemed to be a natural progression from our research. Our final product, a Probabilistic Programming
web application, allows a user to code in JavaScript by way of selecting from
a list of suggestions for the user's next line of code. We imagine that this won't only be helpful for new programmers,
but that this system would allow for expert programmers to write code with more ease and efficiency.

@@

\noindent and ${Path}({node}_1, {edge}_1) = {node}_2$, ${Path}({node}_2, {edge}_2) = {node}_3$, ..., ${Path}({node}_d, {edge}_d) = {newnode}_k$, where ${newnode}_k$ is randomly selected from ${newnode}_1, ..., {newnode}_n$ given the probabilities $P({newnode_1}), ..., P({newnode_n})$. From this it can be seen that generating programs is simply a repetitive application of this function, where for the chosen ${newnode}_k$, we can generate each child of ${newnode}_k$ with the following formula. \begin{align*} &\forall {edge} \in {Features}({newnode}_k), \\
& \hspace{3em} {Path}({newnode}_k, {edge}) = f({node}_2, {edge}_2, ..., {node}_{d}, {edge}_{d}, {newnode}_k, {edge}).
\end{align*}

@@

\clearpage

@@

The Guided IDE offers users an environment for building a program entirely using
our model, using program snippets generated at the cursor's context as building
blocks. The editor presents a window for the program text, and a list of
suggestion code snippets. The user can move a cursor through the code window at
positions corresponding to boundaries of program nodes. As the user moves the
cursor, the snippet suggestions are regenerated using the cursor's node path, to
continuously provide suggestions relevant to the current context of the program. When the
user selects a suggestion snippet, its AST is inserted into the program's AST at
the cursor's position in the tree, and the code window is redrawn. The user can
move the cursor within newly inserted blocks, so that the program can grow
deeper. The initial corpus for the IDE's model consists of the JavaScript files
that comprise our project.

@@

Instead of having a movable cursor in the code window, consider stubs in the
program. Each stub would be a marker in place of an AST node. The user would
select a stub and be given choices of snippets to replace it. Selecting a
snippet would move the current snippet into the program in place of the selected
stub.  The generated snippets could themselves have stubs in them, so that the
user may continue to complete stubs and expand their program.  At the level of
variable identifiers, the snippet suggestions would be taken from the computed
list of in-scope variables at the position of the selected stub, unless the
completion was for a variable declaration, in which case the user would be
prompted to name a new variable.

@@

\section{Conclusions}

@@

Just as our approach of modeling a program with a tree-based Markov chain is a
higher order model than a linear Markov chain of the program source code tokens,
there may be applications of probabilistic modeling to higher-order
representations of programs than the AST, such as a \emph{higher-order abstract syntax}, which promises to ``incorporate name binding information in a uniform and language generic way'' \cite{hosa}. Such an application may model the relationship between different
parts of an AST, and handle scoping of variables to model identifiers
abstractly.

With a more sophisticated model of identifiers that could generate programs with
identifiers in meaningful relation to each other, Probabilistic Programming could
be used to generate semantically valid programs, for greater assistive,
expressive, and programmatic power.

\subsection{Future}

We should

@@

The source code for Probabilistic Programming is available online under an open
source license:
\url{https://github.com/RocHack/jschain}.

The web IDE can be found at
\url{https://csug.rochester.edu/u/clehner/jschain/web/}.
